A generalized definition of multidimensional item response theory parameters

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Positions tables better in the main text
The new positioning avoids blanks

---
title: "A generalized definition of multidimensional item response theory parameters"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Body Text: ['First Paragraph', 'Abstract']
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
abstract: |
  Modern measurement theory demands that Multidimensional Item Response Theory
  accurately represents covariate spaces with correlated dimensions and/or
  non-standard variances. As it is also used in applications in the
  non-cognitive domain, they also need to model negatively-keyeed items
  accurately and interpretably. Henceforth we generalize the
  multidimensional discrimination [@reckase_discriminating_1991] and
  difficulty [@reckase_difficulty_1985] parameters to account for those two
  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
  Model, represented in a latent space with an arbitrary basis, and
  then defining such basis to be a concrete geometrical representation of the
  phenomena we intend to measure. This results in three different versions of
  the multidimensional parameters: The original one, based on the item
  parameters solely, one that incorporates the covariance structure of the
  latent space, and a third one that uses the correlation structure instead.
  Following @ackerman_creating_1994, we also provide a geometrical
  representation based on these parameters, and provide applied examples to
  illustrate the pertinence of these generalized parameters to represent the
  measurement properties of the items accurately. We recommend using the
  version that uses the latent covariance structure for describing the
  properties of the parameters, and the one based on the correlation structure
  for graphical representation. Finally, we discuss the implications of this
  generalization for other multidimensional IRT models, and the parallels of our
  results in common factor model theory.
keywords: |
  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
  item discrimination, item difficulty, item vector representation
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

@@

@tezza_modelo_2018 provide us with another example that allows showcasing
the effect of correlated dimensions on negative discrimination parameters.
In their model, dimensions 1 and 4 are estimated to have a correlation of
`r DIMS_CORR`.
With such latent space structure, items that strongly discriminate in any or
both of these two dimensions change drastically their multidimensional
parameters from the agnostic to the covariance-based version
(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
for example, with items 57 and 60. These items also have different-sign
discrimination parameters in dimensions 1 and 4, so their situation is
similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
Figure \@ref(fig:item-plot-out),
thus the decrease in their $`r MDISC_ITEM`$ parameter.
The effect on the direction is also patent in the angles these items make with
these dimensions, which change up to almost 20º, compared with at most 7º in the
other two dimensions, and around 1º in most cases.

<!---BLOCK_LANDSCAPE_START--->

@@

>>>>

---
title: "A generalized definition of multidimensional item response theory parameters"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Body Text: ['First Paragraph', 'Abstract']
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
abstract: |
  Modern measurement theory demands that Multidimensional Item Response Theory
  accurately represents covariate spaces with correlated dimensions and/or
  non-standard variances. As it is also used in applications in the
  non-cognitive domain, they also need to model negatively-keyed items
  accurately and interpretably. Henceforth we generalize the
  multidimensional discrimination [@reckase_discriminating_1991] and
  difficulty [@reckase_difficulty_1985] parameters to account for those two
  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
  Model, represented in a latent space with an arbitrary basis, and
  then defining such basis to be a concrete geometrical representation of the
  phenomena we intend to measure. This results in three different versions of
  the multidimensional parameters: The original one, based on the item
  parameters solely, one that incorporates the covariance structure of the
  latent space, and a third one that uses the correlation structure instead.
  Following @ackerman_creating_1994, we also provide a geometrical
  representation based on these parameters, and provide applied examples to
  illustrate the pertinence of these generalized parameters to represent the
  measurement properties of the items accurately. We recommend using the
  version that uses the latent covariance structure for describing the
  properties of the parameters, and the one based on the correlation structure
  for graphical representation. Finally, we discuss the implications of this
  generalization for other multidimensional IRT models, and the parallels of our
  results in common factor model theory.
keywords: |
  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
  item discrimination, item difficulty, item vector representation
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

@@

@tezza_modelo_2018 provide us with another example that allows showcasing
the effect of correlated dimensions on negative discrimination parameters.
In their model, dimensions 1 and 4 are estimated to have a correlation of
`r DIMS_CORR`.
With such latent space structure, items that strongly discriminate in any or
both of these two dimensions change drastically their multidimensional
parameters from the agnostic to the covariance-based version
(see Table \@ref(tab:empirical-table-Tezza2018-out))
<!---BLOCK_LANDSCAPE_START--->

@@

---this happens,
for example, with items 57 and 60. These items also have different-sign
discrimination parameters in dimensions 1 and 4, so their situation is
similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
Figure \@ref(fig:item-plot-out),
thus the decrease in their $`r MDISC_ITEM`$ parameter.
The effect on the direction is also patent in the angles these items make with
these dimensions, which change up to almost 20º, compared with at most 7º in the
other two dimensions, and around 1º in most cases.
